Modeling the Linux page cache for accurate simulation of data-intensive applications

### 2020-12-15 paper_io_sim.tex
edits in intro

    The emergence of Big Data in recent years has resulted in a growing
    need for efficient data processing solutions. While infrastructures
    with sufficient compute power are available,
    the I/O bottleneck remains. The Linux page cache is an efficient
    approach to reduce I/O overheads, but few
    experimental studies of its interactions with Big Data applications exist,
    partly due to limitations associated with
    real-world experiments. Simulation is a popular approach to address
    these issues, however, existing simulation frameworks do not simulate
    page caching fully, or even at all.  As a result, simulation-based
    performance studies of data-intensive applications leads to inaccurate
    results.

    In this paper, we propose an I/O simulation model that includes
    the key features of the page cache. We have implemented this model
    as part of the \wrench workflow simulation framework, which itself
    builds on the popular \simgrid distributed systems simulation
    framework. Our model and its implementation enable the simulation
    of both single-threaded and multithreaded applications, and of both
    writeback and writethrough caches for local or network-based
    filesystems. We evaluate the accuracy of our model in different
    conditions, including sequential and concurrent applications, as
    well as local and remote I/Os. We find that our page cache model
    reduces the simulation error by up to an order of magnitude when
    compared to state-of-the-art, cacheless simulations.

@@

        The Linux page cache plays an important role in
        reducing filesystem data transfer times. With the page cache, previously
        read data can be re-read directly from memory and written data can be written to
        memory before being asynchronously flushed to disk, resulting in improved I/O performance
        on slower storage devices. The performance improvements
        depend on many factors including the total amount of memory,
        the amount of data being written (i.e., dirty data), and the amount of memory available for
        written data. All these factors are important when determining the impact of I/O on
        application performance, particularly in data-intensive applications.

        % Swapped the order the two paragraphs and merged them into one
        The number of data-intensive applications has been steadily rising as a result of
        open-data and data sharing initiatives. Due to the sheer size of the data being
        processed, these applications must be executed on large-scale infrastructures
    such as High Performance Computing (HPC) clusters or the cloud.  It
    is thus crucial to quantify the performance of these applications
    on these platforms. The goals include determining which type of hardware/software
    stacks are best suited to different application classes, as well as
    understanding the shortcomings of current algorithms, designs and
    technologies so as to propose new solutions. Unfortunately, application performance
        studies via experiments on real-world compute platforms
        face several difficulties (high operational costs, labor-intensive experimental setups,
        shared platforms with dynamic loads that hinder reproducibility of results) and shortcomings
        (experiments are limited to the available platform/software configurations, which precludes
        the exploration of hypothetical scenarios). 
    Simulations address these concerns by providing models
    and abstractions for the performance of computer hardware, such as
    CPU, network and storage. As a result, simulations provide a
    cost-effective, fast, easy and reproducible way to evaluate
    application performance on arbitrary platform configurations. It thus comes
        as no surprise that, considering here only the field of parallel and distributed
        computing, a large number of simulation frameworks have been developed and used
        to support research and development endeavors~\cite{ optorsim, gridsim, groudsim, cloudsim, nunez2012simcan,nunez2012icancloud, mdcsim, dissect_cf, cloudnetsimplusplus, fognetsimplusplus, casanova2014simgrid, ROSS, casanova2020fgcs}.

        Page caching is an ubiquitous technique for mitigating the I/O bottleneck
        faced by data-intensive applications. As such, it is necessary to simulate page caching
        when performing performance studies in simulation.
        Furthermore, simulating the page cache makes it possible to study relevant application-I/O
        bottlenecks that can arise over the course of execution, such as page cache
        starvation~\cite{zhuang2017}.
        While existing simulation frameworks of parallel and distributed computing
        systems  capture many relevant features of hardware/software stacks, 
        they lack the ability to simulate page cache adequately. Those frameworks
        that implement some page cache simulation
        fail to capture key features such
        as dirty data and cache eviction policies~\cite{nunez2012simcan,nunez2012icancloud}. 
        Some simulators, such as that in~\cite{xu2018saving}, do capture such features
        but are domain-specific.

@@

    In this work, we present \wrench-cache, a simulation model of I/O
    with page cache incorporated into the \wrench simulation framework~\cite{casanova2020fgcs},
    which is  based on the popular \simgrid distributed simulation framework~\cite{casanova2014simgrid}.
    The added value of \wrench-cache is to make it possible to simulate 
        application execution accurately in the presence of page caching. Our specific 
        contributions are as follows:
    \begin{compactitem}
        \item A page cache simulation model that supports 
    both single-threaded and multithreaded applications, and of both
    writeback and writethrough caches for local or network-based
    filesystems;
        \item An implementation of this model in \wrench; 
        \item An evaluation of the accuracy and scalability of our model, and of its implementation,
              for a range of applications, platforms, and page cache configurations. 
    \end{compactitem}

@@

    Page cache simulation would be instrumental in a number of studies. For
    instance, it is now common for HPC clusters to run applications in
    Linux control groups (cgroups), where resource consumption is limited, including
    memory and therefore page cache usage. Using our simulator,
    it would be possible to study the interaction between memory allocation
    and I/O performance, for instance to improve scheduling algorithms.
    Our simulator could also be utilized to evaluate solutions that
    reduce the impact of network file transfers on distributed
    applications, such as burst
    buffers~\cite{ferreiradasilva-fgcs-bb-2019}, hierarchical file
    systems~\cite{islam2015triple}, active storage~\cite{5496981}, or specific
    hardware architectures~\cite{hayot2020performance}.

>>>>

    The emergence of Big Data in recent years has resulted in a growing
    need for efficient data processing solutions. While infrastructures
    with sufficient compute power are available,
    the I/O bottleneck remains. The Linux page cache is an efficient
    approach to reduce I/O overheads, but few
    experimental studies of its interactions with Big Data applications exist,
    partly due to limitations of
    real-world experiments. Simulation is a popular approach to address
    these issues, however, existing simulation frameworks do not simulate
    page caching fully, or even at all.  As a result, simulation-based
    performance studies of data-intensive applications leads to inaccurate
    results.

    In this paper, we propose an I/O simulation model that includes
    the key features of the Linux page cache. We have implemented this model
    as part of the \wrench workflow simulation framework, which itself
    builds on the popular \simgrid distributed systems simulation
    framework. Our model and its implementation enable the simulation
    of both single-threaded and multithreaded applications, and of both
    writeback and writethrough caches for local or network-based
    filesystems. We evaluate the accuracy of our model in different
    conditions, including sequential and concurrent applications, as
    well as local and remote I/Os. We find that our page cache model
    reduces the simulation error by up to an order of magnitude when
    compared to state-of-the-art, cacheless simulations.

@@

        The Linux page cache plays an important role in
        reducing filesystem data transfer times. With the page cache, previously
        read data can be re-read directly from memory, and written data can be written to
        memory before being asynchronously flushed to disk, resulting in improved I/O performance
        on slower storage devices. The performance improvements
        depend on many factors including the total amount of memory,
        the amount of data being written (i.e., dirty data), and the amount of memory available for
        written data. All these factors are important when determining the impact of I/O on
        application performance, particularly in data-intensive applications.

        % Swapped the order the two paragraphs and merged them into one
        The number of data-intensive applications has been steadily rising as a result of
        open-data and data sharing initiatives. Due to the sheer size of the data being
        processed, these applications must be executed on large-scale infrastructures
    such as High Performance Computing (HPC) clusters or the cloud.  It
    is thus crucial to quantify the performance of these applications
    on these platforms. The goals include determining which type of hardware/software
    stacks are best suited to different application classes, as well as
    understanding the limitations of current algorithms, designs and
    technologies. Unfortunately, performance
        studies relying on real-world experiments on compute platforms
        face several difficulties (high operational costs, labor-intensive experimental setups,
        shared platforms with dynamic loads that hinder reproducibility of results) and shortcomings
        (experiments are limited to the available platform/software configurations, which precludes
        the exploration of hypothetical scenarios). 
    Simulations address these concerns by providing models
    and abstractions for the performance of computer hardware, such as
    CPU, network and storage. As a result, simulations provide a
    cost-effective, fast, easy and reproducible way to evaluate
    application performance on arbitrary platform configurations. It thus comes
        as no surprise that a large number of simulation frameworks have been developed and used 
        for research and development~\cite{ optorsim, gridsim, groudsim, cloudsim, nunez2012simcan,nunez2012icancloud, mdcsim, dissect_cf, cloudnetsimplusplus, fognetsimplusplus, casanova2014simgrid, ROSS, casanova2020fgcs}.

        Page caching is an ubiquitous technique for mitigating the I/O bottleneck.
        As such, it is necessary to model it
        when simulating data-intensive applications.
        While existing simulation frameworks of parallel and distributed computing
        systems  capture many relevant features of hardware/software stacks, 
        they lack the ability to simulate page cache with enough details to capture key features such
        as dirty data and cache eviction policies~\cite{nunez2012simcan,nunez2012icancloud}. 
        Some simulators, such as the one in~\cite{xu2018saving}, do capture such features,
        but are domain-specific.

@@

    In this work, we present \wrench-cache, a page cache simulation model
    implemented in \wrench~\cite{casanova2020fgcs}, a workflow simulation
    framework based on the popular \simgrid distributed simulation
    toolkit~\cite{casanova2014simgrid}. Our contributions are:
    \begin{compactitem}
        \item A page cache simulation model that supports 
    both single-threaded and multithreaded applications, and of both
    writeback and writethrough caches for local or network-based
    filesystems;
        \item An implementation of this model in \wrench; 
        \item An evaluation of the accuracy and scalability of our model, and of its implementation,
              for multiple applications, execution scenarios, and page cache configurations. 
    \end{compactitem}

@@

    Page cache simulation would be instrumental in a number of studies. For
    instance, it is now common for HPC clusters to run applications in
    Linux control groups (cgroups), where resource consumption is limited,
    including memory and therefore page cache usage. Using our simulator,
    it would be possible to study the interaction between memory allocation
    and I/O performance, for instance to improve scheduling algorithms or
    avoid page cache starvation~\cite{zhuang2017}. Our simulator could also
    be utilized to evaluate solutions that reduce the impact of network
    file transfers on distributed applications, such as burst
    buffers~\cite{ferreiradasilva-fgcs-bb-2019}, hierarchical file
    systems~\cite{islam2015triple}, active storage~\cite{5496981}, or
    specific hardware architectures~\cite{hayot2020performance}.
